A decision support tool for the Home Health Care Routing and Scheduling Problem

### 2021-12-18 tex/Paper/main.tex
updated files.

\begin{abstract}
The Home Health Care Routing and Scheduling Problem (HHCRSP) consists in assigning routes and schedules to nurses that deliver health care services to patients at their homes. A varied range of hard and soft constraints such as time windows, synchronisation of nurses, skill mix or patient and nurse preferences make this problem particularly difficult to solve manually. However, this is a challenge faced daily by many community nursing teams.

@@

We compare the performance of the GRASP algorithm with state-of-the-art approaches for similar problems and show that it offers competitive results, while providing a wider range of possibilities when modelling real problems. For instance, the computational results show that considering waiting times in the objective function can reduce total operation times up to 19\% in comparison with alternative modelling approaches that minimise travel time only.

@@

A key role in the health service to deliver this care are District Nurses (DN), specialised nursing teams that travel to patient's homes and perform a variety of tasks, ranging from insulin injections to dressing wounds. Nevertheless, DN numbers have fallen by 44\% \citep{nhs2018} between 2010 and 2018 in the NHS. On the other hand, while there is no official data on demand, the workforce perception is that demand has risen significantly, putting the service under pressure \citep{robertson2017}. There is, therefore, a strong need for efficiency within the community nursing service, to better utilise the available resources.

One of the first tasks that the DN team lead faces on a working day is to distribute the work for the team. In the DN teams we visited in the city of Southampton, UK, this entailed assigning a number of home health care visits (typically over a hundred) to the nurses available on the team, attending to hard requirements (skill matching, time windows, job dependencies) and soft ones (minimising travel and waiting times, maintaining workload balance, matching carer-patient preferences). Despite the abundant literature related to similar problems, the planning is currently performed manually in a lengthy process for the planner that potentially leads to suboptimal schedules. Manual planning has three further weaknesses. First there are inconsistencies between the different plans of a nurse, and between plans of different nurses. Although a single DN may become highly skilled in planning overtime that knowledge is lost when they leave the organisation. Second, a poor planning results in turn in a poor utilisation of the workforce, which might include excessive travel and waiting times, unbalanced workloads. Third, it can also affect the quality of care, leading to missed time windows or lack of preference matching for patients.

In an effort to automate and improve this process, we model the planning of home health care visits as a rich vehicle routing problem, with constraints and objectives drawn from our interaction with community nursing teams. This problem is often known in the literature as Home Health Care Routing and Scheduling Problem (HHCRSP). In our case, the schedule takes place over a single day, as it is the preferred approach in practice. This is motivated by the last-minute changes due to staff availability, patient next-day referrals and the nature of continuous review of care, which often prompts to scheduling the next visit right after reviewing their condition on the previous visit. A formal description of the problem is provided in Section~\ref{sec:model}.

We propose a Decision Support Tool (DST) that can produce high quality daily routes and schedules for a DN team, according to a flexible objective function. The tool requires as input a patient requirement list and the available staff and then runs an optimisation algorithm based on a weighted objective function. The optimisation is based on a Greedy Randomised Adaptive Search Procedure (GRASP) with Path Relinking, which is detailed in Section~\ref{sec:dst}.

The main contribution of this article is to present a DST for the HHCRSP that is designed to be used in real community nursing settings. It tackles the most common constraints found in literature (e.g. time windows or skills) but also others often found in practice (e.g. multiple depots or workload balance). We provide extensive computational results (Section~\ref{sec:computationalexperiments}) that show that our algorithm is competitive with state-of-the-art approaches and is flexible to allow real-world constraints and objectives; but also provide insight on the algorithm components that help achieve high quality routes and schedules.

%----------------------------------------------------------------------------------%
\section{Literature review}
\label{sec:litreview}
\noindent The generalised growth in demand and availability of home health care services of recent years has resulted in a growth in academic interest in the related scheduling problem. While the problem might be addressed by drawing from the extensive vehicle routing problem (VRP) literature (see, for example, \cite{drexl2012} for a comprehensive review of VRPs and synchronisation constraints), HHCRSP has developed its own body of literature. In particular, there are recent review articles: \cite{cisse2017}, \cite{fikar2017}, \citet{dimascolo2021}, and \citet{grieco2020}. We refer the reader to them for a broader understanding of what is meant by HHCRSP and different solving approaches. Home Health Care and Home Care services across the world operate in different ways and, as such, the related scheduling problems have different objectives and constraints. For example, it might be possible to speed up the visits to some patients \citep{mosquera2019}, to allow nurses to share a minibus for part of their route, while walking the rest of it \citep{fikar2015}, for nurses to use public transport \citep{rest2016}, or to allow time preferences for patients \citep{decerle2018}. % This variety makes difficult comparing approaches and translating the findings from one application to another.

@@

% Exact methods
Several articles propose exact methods. These are usually based on a Mixed Integer Linear Program (MILP) formulation that is solved with the help of a commercial solver \citep{bredstrom2008,bachouch2011, mankowska2014, aithaddadene2016}. Typically, these formulations are three index formulations that involve two types of variables: $x_{ijk}$ binary variables that indicate if carer $k$ traverses the arc $i$ and $j$; and $t_{ik}$ real variables that determine at what time carer $k$ starts performing job $i$. These formulations are limited in practice and can only solve to optimality small problems (e.g. 45 patients in \cite{aithaddadene2016} or 10 in \cite{mankowska2014}). In order to address this, some researchers propose to decompose the problem in a column generation scheme. For example, \cite{rasmussen2012} tackles the HHCRSP problem in Denmark. They propose a branch-and-price scheme where generalised synchronisation constraints are relaxed and dealt with in the branching.% \correction{Only Rasmussen?}
 % \correction{Cite relevant papers here}. The ILP formulations work well for small problems, but are often not practical to solve for problems with a significant number of patients. %around 50

Often, in order to tackle larger instances, the MILPs are solved in a restricted manner and, while they do not guarantee optimality, they might provide feasible solutions suitable for day to day planning. This is often referred to as a matheuristic algorithm and can be done by reducing the size of the original problem by fixing some aspects and solving it again. One example of this technique is given in \cite{rasmussen2012} that also proposes clustering strategies to reduce the size of the problem, which allows them to solve problems with up to $150$ patients, although no longer proving optimality. Another is \cite{bredstrom2008}, where they propose a heuristic that iteratively allows or forbids assignments between patients and carers in order to generate a series of restricted MILPs that can be solved in a short time. In addition, \citet{nikzad2021} considers a four-phase matheuristic that calls upon a variety of mixed integer models.

@@

Finally, we look at the literature that discusses the software implementation of the solution approaches in practice. One of the earliest works is due to \cite{begur1997}, who developed a decision support system based on the Clarke and Wright algorithm \citep{clarke1964} for vehicle routing and then relied on integration with a Geographic Information System (GIS), an interactive user interface and a human planner to alter the routes as necessary. The Laps Care system \citep{eveborn2006, eveborn2009} is a Swedish software for the HHCRSP, based on a repeated matching algorithm has been shown to be successful in saving planning time, minimising travel times and even reducing the number of missed visits. The authors estimated annual savings in 2009 of up to \euro 120 million across teams in Norway, Finland and Sweden \citep{eveborn2009}. Another software is Parpap \citep{bertels2006}, which proposes various approaches combining constraint programming and metaheuristics. A common claim in all these implementation-oriented works is that the software needs to be flexible enough to allow for real-world constraints to be implemented and offers interaction with mapping software.

%----------------------------------------------------------------------------------%
\section{A model for health home care services}
\label{sec:model}
\noindent In this section we describe the notation and give a formal description of the HHCRSP problem tackled by our DST.

We consider a list of nurses $N = \{1,\dots,n\}$ that are available for the day. Each nurse $i$ might have a different starting time, $st(i)$, and end time, $en(i)$, of their working day. During the planning horizon, which is one day in our case, the nurses need to perform a number of jobs $J = \{1,\dots,m\}$, a subset of which, $DS \subset J$ are double services, i.e. they require two nurses to be present to be performed. Note that jobs refer to individual tasks that a nurse needs to perform and are not necessarily identified with unique patients or locations, meaning that the same patient might have more than one job on the list.

A schedule (not necessarily feasible) $\mathcal{S}$, is made of a pair   $\mathcal{S} = \{\mathcal{R}, \mathcal{O}\}$, where $\mathcal{O}$ is a permutation of the vector $o = (1,\dots,n)$ representing the order of the nurses in the solution and $\mathcal{R}$ is a set of $n$ routes, $\mathcal{R} = \{r_1,\dots,r_n\}$, one per nurse. Each route $r_i$ is itself an ordered list of $m_i$ jobs, $r_i = \{k^i_0, k^i_1,\dots,k^i_{m_i}, k^i_0\}$, that are the jobs assigned to nurse $i \in N$ and starts and ends at the depot of that nurse denoted by, $k^i_0$. Of course, a route might be empty, meaning that that particular nurse is not assigned any jobs and $m_i = 0$ and the route only contains the depot, $r_i = \{k^i_0, k^i_0\}$.
After a schedule is generated, for each job $j \in J$ we denote by $t_j$ its starting time. Note that this refers to the time at which the job starts, rather than the time at which the nurse (or nurses) performing it arrive to the job location, which might be different.

In the rest of the section, we formally define the constraints and objectives that commonly apply in practice and relate to the skill mix of the workforce, the preferences of some assignments, travel times and synchronisation, visits with two staff members and quality of the schedule.

%----------------------------------------------------------------------------------%
\subsection{Skill Mix}
\label{sub:skillmix}
\noindent Some jobs might require a specialised nurse to be performed, and nurses might only perform jobs that they are qualified to do. To model this, we define a function $S_1: N\times J \rightarrow \{0,1\}$, that takes value $1$ if nurse $i$ is skilled to perform job $j$, $S_1(i,j) = 1$ and $0$ otherwise. 
Naturally, for double service jobs, no single nurse is skilled to perform them $S_1(i,j) = 0, \forall i \in N, \forall j \in DS$. To account for this, we define a second function $S_2: N\times N\times J \rightarrow \{0,1\}$, that takes value $S_2(i_1,i_2,j) = 1$ if the combination of nurses $i_1$ and $i_2$ is sufficiently skilled to perform job $j$ and $0$ otherwise.

@@

A relevant modelling consideration relates to the necessary skill mix for simultaneous double services or, in other words, how the function $S_2$ is created in practice. We distinguish the following cases:
\begin{itemize}
    \item \textbf{Duplicated}. In this case it is required for all staff members visiting the patient to have all necessary skills to perform the service.
    \item \textbf{Shared}. This is the case when it is sufficient that the combination of skills between the staff members present is enough to perform the service. In particular, it could happen that only one staff member is qualified for all services and the other for none. An example of this would be visits that require a second person to provide support and ensure safety, for example with patients that have a record of behaving inappropriately towards staff.
    \item \textbf{Strictly shared}. In this case each member of the service does not need to be qualified for all services but should be attending the visit to perform at least one service. This is the approach used in \cite{mankowska2014} and \cite{aithaddadene2016}.
    \item \textbf{Shared-Duplicated}. This case is analogous to \emph{Shared}, but it allows some skills to be required by more than one of the staff members, hence it generalises the previous cases.
\end{itemize}

\noindent In our DST, the default approach is \emph{Shared-Duplicated}, as it allows for the flexibility that is observed in real operations. The data is introduced as a list of skills and, if some skills need to be present in more than one staff member, they can simply appear duplicated on the list. We have, however, also implemented the Strictly Shared approach, in order to solve the instances from the literature.

It is not uncommon in the literature to model the skills of the workforce as hierarchical, i.e. each nurse is assigned a skill level and it is assumed that they possess all the skills that lower levels have, and some more. The model might then introduce a penalty or limit for high skilled nurses performing too many low skilled jobs, known as downgrading \cite[see, for example,][]{fikar2015}. In the DN teams we observed, skills were not hierarchical, neither were any such limits in place. Nevertheless, nurses pointed out that high skilled nurses were typically assigned only the high skilled jobs, as they might need to perform other jobs (e.g. attending meetings or other administrative tasks). While this is not something that we consider explicitly, a similar objective may be achieved through setting negative preferences between high skilled nurses and low skilled jobs.

%----------------------------------------------------------------------------------%
\subsection{Preferences}
\label{sub:preferences}
\noindent It is not uncommon in practice that district nurses do their planning using preferences. These might arise from multiple reasons, such as a preserving continuity of care, gender preference, patients and nurses speaking the same language, previous conflicts between a patient and a staff member, \model{etc}. We model this with a preference function, $P: N\times J \rightarrow \mathbb{R}$, such that $P(i,j)$ represents the preference score of assigning job $i$ to staff member $j$. The function $P$ should assign a higher (positive) score to preferred patient-nurse assignments and a lower (negative) score to assignments that are better avoided. We give an indication on how to choose the specific values of the function in Section~\ref{subsub:adequateweights}.

%----------------------------------------------------------------------------------%
\subsection{Travel Times}
\label{sub:traveltimes}
\noindent To describe travel times, for each nurse we define $N$ functions $T_i: J\cup\{k^i_0\} \times J\cup\{k^i_0\} \rightarrow \mathbb{R}$, $i \in N$, where $T_i(j_1, j_2)$ indicates the time it takes nurse $i$ to travel from job $j_1$ to job $j_2$, where $k^i_0$ refers to the start and end location of the nurse $i$. Each nurse might start from a different depot (typically, their own home), a fact that is accounted for in these functions. Note that with this definition we allow to have different travel times between jobs for different nurses, and this can be used to model, for example, different means of transport. In our observations, however, all nurses used their own vehicles, so we assume that $T_i$ returns driving times for all nurses and, in particular, that:
\begin{equation}
    T_{i_1}(j_1, j_2) = T_{i_2}(j_1, j_2),\ \forall i_1, i_2 \in N,\ \forall j_1, j_2 \in J. \label{eqn:traveltimequality}
\end{equation}

\noindent Note that this function can be substituted by a more sophisticated one without any changes to the algorithms described here.

@@

%----------------------------------------------------------------------------------%
\subsection{Temporal Constraints}
\label{sub:temporalconstraints}
\noindent Each job $j$ has a service time $s_j$ and a time window $t_j$ that is defined by its earliest start time, $e_j$ and the latest start time $l_j$. It is a hard constraint that jobs cannot start before their earliest start time:
\begin{align}
    t_j \geq e_j, & \forall j \in J, \label{eqn:jobstart}
\end{align}

\noindent however, they may start late. The tardiness with which a job $j$ starts, $z_j$, can be simply calculated using the following:
\begin{equation}
    z_j = \max(0, t_j - l_j). \label{eqn:tardiness}
\end{equation}

\noindent Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses might be further restricted temporally. This might happen in practice, for example, in situations where a certain medication must be administered \model{a few hours (a specific time period/amount of time)} after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$ that depends on the start time of $j_1$, $[e^{(j_1)}_{j_2} + t_{j_1}, l^{(j_1)}_{j_2} + t_{j_1}]$. Then, the same time window constraints hold:
\begin{subequations}
	\begin{align}
	t_{j_2} &\geq e^{(j_1)}_{j_2} + t_{j_1}, \label{eqn:timewindow1}\\[3pt] % & \forall j \in J
	t_{j_2} &\leq l^{(j_1)}_{j_2} + t_{j_1}. \label{eqn:timewindow2} % & \forall j \in J
	\end{align}
\end{subequations}

\noindent Similarly, $j_2$ will impose a time window on $j_1$, and in a valid instance of the problem it must hold that:
\begin{subequations}
	\begin{align}
	e^{(j_1)}_{j_2} &= -e^{(j_2)}_{j_1}, \label{eqn:earliesttimedependecies}\\[3pt]
	l^{(j_1)}_{j_2} &= -l^{(j_2)}_{j_1}. \label{eqn:latesttimedependencies}
	\end{align}
\end{subequations}

\noindent Then, we can define the \emph{gap tardiness} of job $j_2$, $\tilde{z}_{j_2}$ as:
\begin{equation}
    \tilde{z}_{j_2} = \max(0, t_{j_2} - (l^{(j_1)}_{j_2} + t_{j_1})). \label{eqn:gaptardiness}
\end{equation}

\noindent \model{Since some jobs have time windows (only \textit{some}?)} or might need to be synchronised, waiting times may occur. We denote by $w_{ij}$ the waiting time of nurse $i$ at job $j$. Note that $w_{ij} = 0$ if $j \notin r_i$; otherwise, $w_{ij}$ represents the difference between $t_j$ and the arrival time of nurse $i$ to job $j$.% , denoted by $a(i,j)$.

The arrival time $a(i,j)$ of nurse $i$ at job $j \in r_i$ is calculated as follows:
\begin{equation}
    a(i,j) = st(i) + \sum_{h=0}^{j' - 1} w_{ik^i_h} + s_{k^i_h} + T_i(k^i_h, k^i_{h + 1}),
\end{equation}

\noindent where $j'$ denotes the position of job $j$ in the route $r_i$, i.e. $j = k^i_{j'} \in r_i$. Note that there are no waiting or service times at the depot; that is, $w_{ik^i_0} = 0$ and $s_{k^i_0} = 0 $ for all $i \in N$.

In practice, waiting times may occur for a variety of reasons. In order to maintain simplicity in our model, we only consider three reasons for which a nurse $i$ might have waiting time $w_{ij} \geq 0$ at a job $j \in r_i$:
\begin{itemize}
    \item Respecting a time window: 
    \begin{align}
        w_{ij} \geq e_i - a(i,j) & \forall i \in N, \forall j \in J.
    \end{align}
    \item Waiting for a second nurse $i'$ to arrive for a double service:
    \begin{align}
        w_{i j} \geq t_j - a(i',j) & \forall j \in DS,\ such\ that\ j \in r_{i} \cap r_{i'}.
    \end{align}
    \item Waiting for the minimum gap of a synchronised service. If the job $j$ depends on $j'$, then:
    \begin{equation}\label{eqn:waitingforsynchronised}
        w_{ij} \geq e^{(j')}_{j} - a(i,j).
    \end{equation}
\end{itemize}

>>>>

\author[1]{Asyl L. Hawa}
\ead{A.Hawa@soton.ac.uk}

\begin{abstract}
The Home Health Care Routing and Scheduling Problem (HHCRSP) consists in assigning routes and schedules to nurses that deliver health care services to patients at their homes. A varied range of hard and soft constraints such as time windows, synchronisation of nurses, skill mix, and patient and nurse preferences make this problem particularly difficult to solve manually. However, this is a challenge faced daily by many community nursing teams.

@@

We compare the performance of the GRASP algorithm with state-of-the-art approaches for similar problems and show that it offers competitive results, while providing a wider range of possibilities when modelling real problems. For instance, the computational results show that considering waiting times in the objective function can reduce total operation times up to 24\% in comparison with alternative modelling approaches that minimise travel time only.

@@

A key role in the health service to deliver this care are District Nurses (DN), specialised nursing teams that travel to patients' homes and perform a variety of tasks, ranging from insulin injections to dressing wounds. Nevertheless, DN numbers have fallen by 44\% \citep{nhs2018} between 2010 and 2018 in the NHS. On the other hand, while there is no official data on demand, the workforce perception is that demand has risen significantly, putting the service under pressure \citep{robertson2017}. There is, therefore, a strong need for efficiency within the community nursing service to better utilise the available resources.

One of the first tasks that the DN team lead faces on a working day is to distribute the work for the team. In the DN teams we visited in the city of Southampton, UK, this entailed assigning a number of home health care visits (typically over a hundred) to the nurses available on the team, attending to hard requirements (skill matching, time windows, job dependencies) and soft ones (minimising travel and waiting times, maintaining workload balance, matching nurse-patient preferences). Despite the abundant literature related to similar problems, the planning is currently performed manually in a lengthy process for the planner that potentially leads to suboptimal schedules. Manual planning has three further weaknesses. First, there are inconsistencies between the different plans of a nurse, and between plans of different nurses. Although a single DN may become highly skilled in planning over time, that knowledge is lost when they leave the organisation. Second, poor planning results in poor utilisation of the workforce, which may include excessive travel and waiting times, and unbalanced workloads. Third, maunal planning can affect the quality of care, leading to missed time windows or lack of preference matching for patients.

In an effort to automate and improve this process, we model the planning of home health care visits as a rich vehicle routing problem with constraints and objectives drawn from our interaction with community nursing teams. This problem is often known in the literature as the Home Health Care Routing and Scheduling Problem (HHCRSP). In our case, the schedule takes place over a single day, as it is the preferred approach in practice. This is motivated by the last-minute changes due to staff availability, patient next-day referrals and the nature of continuous review of care, which often prompts to scheduling the next visit right after reviewing the patient's condition on the previous visit. A formal description of the problem is provided in Section~\ref{sec:model}.

We propose a Decision Support Tool (DST) that can produce high quality daily routes and schedules for a DN team, according to a flexible objective function. The tool requires as input a patient requirement list and the available staff, and then runs an optimisation algorithm based on a weighted objective function. The optimisation is based on a Greedy Randomised Adaptive Search Procedure (GRASP) with Path Relinking, which is detailed in Section~\ref{sec:dst}.

The main contribution of this article is to present a DST for the HHCRSP that is designed to be used in real community nursing settings. It tackles the most common constraints found in literature (e.g. time windows and skills) but also others often found in practice (e.g. multiple depots and workload balance). We provide extensive computational results (Section~\ref{sec:computationalexperiments}) that show that our algorithm is competitive with state-of-the-art approaches and is flexible to allow real-world constraints and objectives; but also provide insight on the algorithm components that help achieve high quality routes and schedules.

%----------------------------------------------------------------------------------%
\section{Literature Review}
\label{sec:litreview}
\noindent The generalised growth in demand and availability of home health care services of recent years has resulted in a growth in academic interest in the related scheduling problem. While the problem might be addressed by drawing from the extensive vehicle routing problem (VRP) literature (see, for example, \cite{drexl2012} for a comprehensive review of VRPs and synchronisation constraints), HHCRSP has developed its own body of literature. In particular, there are recent review articles: \cite{cisse2017}, \cite{fikar2017}, \citet{dimascolo2021}, and \citet{grieco2020}. We refer the reader to them for a broader understanding of what is meant by HHCRSP and different solving approaches. Home Health Care and Home Care services across the world operate in different ways and, as such, the related scheduling problems have different objectives and constraints. For example, it might be possible to speed up the visits to some patients \citep{mosquera2019}, to allow nurses to share a minibus for part of their route, whilst walking the rest of it \citep{fikar2015}, for nurses to use public transport \citep{rest2016}, or to allow time preferences for patients \citep{decerle2018}. % This variety makes difficult comparing approaches and translating the findings from one application to another.

@@

% Exact methods
Several articles propose exact methods. These are usually based on a Mixed Integer Linear Program (MILP) formulation that is solved with the help of a commercial solver \citep{bredstrom2008,bachouch2011, mankowska2014, aithaddadene2016}. Typically, these formulations are three index formulations that involve two types of variables: $x_{ijk}$ binary variables that indicate if nurse $k$ traverses the arc $i$ and $j$; and $t_{ik}$ real variables that determine at what time nurse $k$ starts performing job $i$. These formulations are limited in practice and can only solve to optimality small problems (e.g. 45 patients in \cite{aithaddadene2016} or 10 in \cite{mankowska2014}). In order to address this, some researchers propose to decompose the problem in a column generation scheme. For example, \cite{rasmussen2012} tackles the HHCRSP problem in Denmark. They propose a branch-and-price scheme where generalised synchronisation constraints are relaxed and dealt with in the branching.% \correction{Only Rasmussen?}
 % \correction{Cite relevant papers here}. The ILP formulations work well for small problems, but are often not practical to solve for problems with a significant number of patients. %around 50

Often, in order to tackle larger instances, the MILPs are solved in a restricted manner and, while they do not guarantee optimality, they might provide feasible solutions suitable for day-to-day planning. This is often referred to as a matheuristic algorithm, and can be done by reducing the size of the original problem by fixing some aspects and solving it again. One example of this technique is given in \cite{rasmussen2012} that also proposes clustering strategies to reduce the size of the problem, which allows them to solve problems with up to $150$ patients, although no longer proving optimality. Another is \cite{bredstrom2008}, where they propose a heuristic that iteratively allows or forbids assignments between patients and nurses in order to generate a series of restricted MILPs that can be solved in a short time. In addition, \citet{nikzad2021} considers a four-phase matheuristic that calls upon a variety of mixed integer models.

@@

Finally, we look at the literature that discusses the software implementation of the solution approaches in practice. One of the earliest works is due to \cite{begur1997}, who developed a decision support system based on the Clarke and Wright algorithm \citep{clarke1964} for vehicle routing and then relied on integration with a Geographic Information System (GIS), an interactive user interface and a human planner to alter the routes as necessary. The Laps Care system \citep{eveborn2006, eveborn2009} is a Swedish software for the HHCRSP based on a repeated matching algorithm, and has been shown to be successful in saving planning time, minimising travel times and even reducing the number of missed visits. The authors estimated annual savings in 2009 of up to \euro 120 million across teams in Norway, Finland and Sweden \citep{eveborn2009}. Another software is Parpap \citep{bertels2006}, which proposes various approaches combining constraint programming and metaheuristics. A common claim in all these implementation-oriented works is that the software needs to be flexible enough to allow for real-world constraints to be implemented and offers interaction with mapping software.

%----------------------------------------------------------------------------------%
\section{A Model for Health Home Care Services}
\label{sec:model}
\noindent In this section we describe the notation and give a formal description of the HHCRSP problem tackled by our DST.

We consider a list of nurses $N = \{1,\dots,n\}$ that are available for the day. Each nurse $i$ might have a different starting time, $st(i)$, and end time, $en(i)$, of their working day. During the planning horizon, which is one day in our case, the nurses need to perform a number of jobs $J = \{1,\dots,m\}$, a subset of which, $DS \subset J$, are double services; that is, they require two nurses to be present simultaneously. Note that jobs refer to individual tasks that a nurse needs to perform and are not necessarily identified with unique patients or locations, meaning that the same patient might have more than one job on the list.

A schedule (not necessarily feasible) $\mathcal{S}$ is made of a pair   $\mathcal{S} = \{\mathcal{R}, \mathcal{O}\}$, where $\mathcal{O}$ is a permutation of the vector $o = (1,\dots,n)$ representing the order of the nurses in the solution, and $\mathcal{R}$ is a set of $n$ routes, $\mathcal{R} = \{r_1,\dots,r_n\}$, one per nurse. Each route $r_i$ is itself an ordered list of $m_i$ jobs, $r_i = \{k^i_0, k^i_1,\dots,k^i_{m_i}, k^i_0\}$, that are the jobs assigned to nurse $i \in N$ and start and end at the depot of that nurse, denoted by $k^i_0$. Of course, a nurse's route may be empty, meaning that the nurse has not been assigned any jobs; thus $m_i = 0$ and the route only contains the depot, $r_i = \{k^i_0, k^i_0\}$. After a schedule is generated, for each job $j \in J$ we denote by $t_j$ its starting time. Note that this refers to the time at which the job starts, rather than the time at which the assigned nurse (or nurses) arrive to the job location, which may differ.

In the rest of this section we formally define the constraints and objectives that commonly apply in practice and relate to the skill mix of the workforce, the preferences of some assignments, travel times and synchronisation, visits with two nurses, and the quality of the schedule.

%----------------------------------------------------------------------------------%
\subsection{Skill Mix}
\label{sub:skillmix}
\noindent Some jobs might require a specialised nurse to be performed, and nurses might only perform jobs that they are qualified to do. To model this, we define a function $S_1: N\times J \rightarrow \{0,1\}$, that takes value $1$ if nurse $i$ is skilled to perform job $j$, $S_1(i,j) = 1$ and $0$ otherwise. 
Naturally, for double service jobs no single nurse is skilled to perform them, $S_1(i,j) = 0$ $\forall i \in N, \forall j \in DS$. To account for this, we define a second function, $S_2: N\times N\times J \rightarrow \{0,1\}$, that takes value $S_2(i_1,i_2,j) = 1$ if the combination of nurses $i_1$ and $i_2$ is sufficiently skilled to perform job $j$ and $0$ otherwise.

@@

A relevant modelling consideration relates to the necessary skill mix for simultaneous double services or, in other words, how the function $S_2$ is created in practice. We distinguish the following cases:
\begin{itemize}
    \item \textbf{Duplicated}. In this case it is required for all staff members visiting the patient to have all necessary skills to perform the service.
    \item \textbf{Shared}. This is the case when it is sufficient that the combination of skills between the staff members present is enough to perform the service. In particular, it could happen that only one staff member is qualified for all services and the other for none. An example of this in practice would be visits that require a second person to provide support and ensure safety, for example with patients that have a record of behaving inappropriately towards staff.
    \item \textbf{Strictly Shared}. In this case each member of the service does not need to be qualified for all services but should be attending the visit to perform at least one service. This is the approach used in \cite{mankowska2014} and \cite{aithaddadene2016}.
    \item \textbf{Shared-Duplicated}. This case is analogous to \emph{Shared}, but it allows some skills to be required by more than one of the staff members, hence it generalises the previous cases.
\end{itemize}

\noindent In our DST, the default approach is \emph{Shared-Duplicated}, as it allows for the flexibility that is observed in real operations. The data is introduced as a list of skills and, if some skills need to be present in more than one staff member, they can simply appear duplicated on the list. We have, however, also implemented the \emph{Strictly Shared} approach in order to solve the instances from the literature.

It is not uncommon in the literature to model the skills of the workforce as hierarchical, i.e. each nurse is assigned a skill level and it is assumed that they possess all the skills that lower levels have and more. The model might then introduce a penalty or limit for high skilled nurses performing too many low skilled jobs, known as downgrading \cite[see, for example,][]{fikar2015}. In the DN teams we observed skills were not hierarchical, neither were any such limits in place. Nevertheless, nurses pointed out that high skilled nurses were typically assigned only the high skilled jobs as they might need to perform other jobs (e.g. attending meetings or other administrative tasks). While this is not something that we consider explicitly, a similar objective may be achieved through setting negative preferences between high skilled nurses and low skilled jobs.

%----------------------------------------------------------------------------------%
\subsection{Preferences}
\label{sub:preferences}
\noindent It is not uncommon in practice for district nurses to do their planning using preferences. These might arise from multiple reasons, such as a preserving continuity of care, gender preference, patients and nurses speaking the same language, or previous conflicts between a patient and a nurse. We model this with a preference function, $P: N\times J \rightarrow \mathbb{R}$, such that $P(i,j)$ represents the preference score of assigning job $i$ to nurse $j$. The function $P$ should assign a higher (positive) score to preferred patient-nurse assignments and a lower (negative) score to assignments that are better avoided. We give an indication on how to choose the specific values of the function in Section~\ref{subsub:adequateweights}.

%----------------------------------------------------------------------------------%
\subsection{Travel Times}
\label{sub:traveltimes}
\noindent To describe travel times, for each nurse we define $N$ functions, $T_i: J\cup\{k^i_0\} \times J\cup\{k^i_0\} \rightarrow \mathbb{R}$, $i \in N$, where $T_i(j_1, j_2)$ indicates the time it takes nurse $i$ to travel from job $j_1$ to job $j_2$, and $k^i_0$ refers to the start and end location of the nurse $i$. Each nurse might start from a different depot (typically their own home), a fact that is accounted for in these functions. Note that with this definition we accommodate for different travel times between jobs for different nurses which can be used to model, for example, different means of transport. In our observations, however, all nurses used their own vehicles; thus we assume $T_i$ returns driving times for all nurses and, in particular, that:
\begin{equation}
    T_{i_1}(j_1, j_2) = T_{i_2}(j_1, j_2)\quad  \forall i_1, i_2 \in N,\ \forall j_1, j_2 \in J. \label{eq:traveltimeequality}
\end{equation}

\noindent Note that this function can be substituted by a more sophisticated one without any changes to the algorithms described in the remainder of the paper.

@@

%----------------------------------------------------------------------------------%
\subsection{Temporal Constraints}
\label{sub:temporalconstraints}
\noindent Each job $j$ has a service time $s_j$ and a time window $t_j$ that is defined by its earliest start time, $e_j$, and the latest start time, $l_j$. It is a hard constraint that jobs cannot start before their earliest start time:
\begin{equation}
    t_j \geq e_j \quad \forall j \in J, \label{eq:jobstart}
\end{equation}

\noindent however, they may start late. The tardiness with which a job $j$ starts, $z_j$, can be simply calculated using the following:
\begin{equation}
    z_j = \max(0, t_j - l_j). \label{eq:tardiness}
\end{equation}

\noindent Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses may be further restricted temporally. This can occur in practice, for example, in situations where a second dose of medication must be administered a few hours after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$ that depends on the start time of $j_1$, $[e^{(j_1)}_{j_2} + t_{j_1}, l^{(j_1)}_{j_2} + t_{j_1}]$. Then, the same time window constraints hold:
\begin{subequations}
	\begin{align}
	t_{j_2} &\geq e^{(j_1)}_{j_2} + t_{j_1}, \label{eq:timewindow1}\\[3pt] % & \forall j \in J
	t_{j_2} &\leq l^{(j_1)}_{j_2} + t_{j_1}. \label{eq:timewindow2} % & \forall j \in J
	\end{align}
\end{subequations}

\noindent Similarly, $j_2$ will impose a time window on $j_1$, and in a valid instance of the problem it must hold that:
\begin{subequations}
	\begin{align}
	e^{(j_1)}_{j_2} &= -e^{(j_2)}_{j_1}, \label{eq:earliesttimedependecies}\\[3pt]
	l^{(j_1)}_{j_2} &= -l^{(j_2)}_{j_1}. \label{eq:latesttimedependencies}
	\end{align}
\end{subequations}

\noindent Then, we can define the \emph{gap tardiness} of job $j_2$, $\tilde{z}_{j_2}$ as:
\begin{equation}
    \tilde{z}_{j_2} = \max(0, t_{j_2} - (l^{(j_1)}_{j_2} + t_{j_1})). \label{eq:gaptardiness}
\end{equation}

\noindent Since some jobs have time windows or might need to be synchronised, waiting times may occur. We denote by $w_{ij}$ the waiting time of nurse $i$ at job $j$. Note that $w_{ij} = 0$ if $j \notin r_i$; otherwise, $w_{ij}$ represents the difference between $t_j$ and the arrival time of nurse $i$ to job $j$.% , denoted by $a(i,j)$.

The arrival time $a(i,j)$ of nurse $i$ at job $j \in r_i$ is calculated as follows:
\begin{equation}
    a(i,j) = st(i) + \sum_{h=0}^{j' - 1} w_{ik^i_h} + s_{k^i_h} + T_i(k^i_h, k^i_{h + 1}), \label{eq:arrivaltime}
\end{equation}

\noindent where $j'$ denotes the position of job $j$ in the route $r_i$, i.e. $j = k^i_{j'} \in r_i$. Note that there are no waiting or service times at the depot; that is, $w_{ik^i_0} = 0$ and $s_{k^i_0} = 0 $ for all $i \in N$. Furthermore, let $en_{\text{act}}(i)$ be the actual end time of a nurse $i \in N$, that is:
\begin{equation}
	en_{\text{act}}(i) = a(i,m_i) + w_{ik_{m_i}^i} + s_{k_{m_i}^i}, \label{eq:actualendtime}
\end{equation}
where $m_i$ is the final job in nurse $i$'s route $r_i$.
